SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-07-07 scipy-1.0/paper.tex
Early draft of tech improvements in scipy.stats in last 3 years.

\pgfplotsset{compat=newest}

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

While it is not surprising that Python is heavily used in SciPy, the
usage distribution of the other languages warrants some discussion. Fortran
is an extremely well-established scientific programming language, both
for historical reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We
wrap the FFTPACK Fortran library for performing Fourier
transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251} since
this library has been a standard in the field for 33 years and has a license
that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial 
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also warp the Fortran libraries QUADPACK\cite{1983qspa.book.....P} (for numerical
integration of one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103} (curve-fitting /
interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal distance regression),
MINPACK\cite{osti_6997568} (minimization of linear and nonlinear equations),
ARPACK\cite{leh:sor:yan96} (solving large scale eigenvalue problems), 
ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling Bessel Functions), and 
CDFLIB\cite{CDFLIB_site} (evaluation of cumulative density functions).

@@

\subsection*{API and ABI evolution}

@@

\subsection*{Statistical distributions}

\subsection*{Polynomial interpolators}

@@

    \subsubsection*{Test suite}
    The SciPy test suite is orchestrated by a continuous integration matrix
    that includes posix and Windows (32/64-bit) platforms managed by Travis CI and appveyor,
    respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and include
    code linting with pyflakes and pycodestyle. There are more than $13000$ unit
    tests in the test suite, which is written for usage with the pytest runner, and
    with a 76.5 \% coverage of approximately $204000$ lines
    of code. Documentation for the code is automatically built and hosted by the
    CircleCI service to facilitate evaluation of documentation changes / integrity.
    Our full test suite also passes with PyPy3, a just-in-time compiled version
    of the Python language.
    % NOTE: is there a citation for PyPy?

@@

>>>>

\pgfplotsset{compat=newest}
\usepackage{listings, textcomp}

\lstset{ %
  basicstyle=\ttfamily\footnotesize,  % size of fonts used for the code
  breaklines=true,   % automatic line breaking only at whitespace
  captionpos=b,   % sets the caption-position to bottom
  commentstyle=\color{gray},  % comment style
  keywordstyle=\color{blue},  % keyword style
  stringstyle=\color{red},  % string literal style
  upquote=true  %straight single quotes (requires textcomp)
}

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\code}[1]{\texttt{#1}}

@@

For implementing new functionality, we have a clear order of language preference.
First Python, if performance is not an issue. If it is, then in order of
preference: Cython, C, C++, Fortran. The main motivation for this is 
maintainability: Cython has the highest abstraction level and most 
Python developers will understand it. C is also widely known,
and easier to deal with for the core development team
than C++ and especially Fortran.

While it is not surprising that Python is heavily used in SciPy, the
usage distribution of the other languages warrants some discussion. Fortran
is an extremely well-established scientific programming language, both
for historical reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We
wrap the FFTPACK Fortran library for performing Fourier
transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251} since
this library has been a standard in the field for 33 years and has a license
that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial 
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also wrap the Fortran libraries QUADPACK\cite{1983qspa.book.....P} (for numerical
integration of one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103} (curve-fitting /
interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal distance regression),
MINPACK\cite{osti_6997568} (minimization of linear and nonlinear equations),
ARPACK\cite{leh:sor:yan96} (solving large scale eigenvalue problems), 
ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling Bessel Functions), and 
CDFLIB\cite{CDFLIB_site} (evaluation of cumulative density functions).

@@

\subsection*{API and ABI evolution}
In general, we encourage changes that improve clarity in the API of the library,
but strongly discourage breaking backwards compatibility, given our position near
the base of the scientific Python computing stack.

@@

\subsection*{Statistical distributions}
Version 1.0 of SciPy saw the addition of the energy distance algorithm in
\texttt{scipy.stats.energy\_distance}. Energy distance is used in a varity of applications ranging
from genetic analysis to multivariate independence. We also added the
first Wasserstein distance algorithm to the library in
\texttt{scipy.stats.wasserstein\_distance}. The Wasserstein distance is often
used in color digital image analysis.

We now also provide the cumulative distribution function of the multivariate
normal distribution, \\\texttt{scipy.stats.multivariate\_normal}, via the newly
associated \texttt{logcdf} and \texttt{cdf} methods.

Other recent additions to \texttt{scipy.stats} include the Argus distribution
in \texttt{scipy.stats.argus} (used in particle physics), a class for
generating a distribution from a histogram in
\texttt{scipy.stats.rv\_histogram}, an implementation of the multinomial
distribution in \texttt{scipy.stats.multinomial} (used for example in natural
language processing), a weighted Kendall's tau statistic algorithm in
\texttt{scipy.stats.weightedtau}, interquantile range handling in
\texttt{scipy.stats.iqr}, the Burr type XII distribution is available in
\texttt{scipy.stats.burr12}, the skew normal distribution has been implemented 
as \texttt{scipy.stats.skewnorm}, a trepzoidal continuous random variable
has been added in \texttt{scipy.stats.trapz}, and the matrix normal distribution
is now available in \texttt{scipy.stats.matrix\_normal}.

From a design standpoint, many functions in \texttt{scipy.stats} now support
a \texttt{nan\_policy} keyword to allow user-level control over the algorithmic
decision to propagate, omit, or raise an error in the case that \texttt{NaN}
values are encountered.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polynomial interpolators}
\input{poly}

@@

    \subsubsection*{Test suite}
    The SciPy test suite is orchestrated by a continuous integration matrix
    that includes posix and Windows (32/64-bit) platforms managed by Travis CI and appveyor,
    respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and include
    code linting with pyflakes and pycodestyle. There are more than $13000$ unit
    tests in the test suite, which is written for usage with the pytest runner, and
    with a 76.5 \% coverage of approximately $204000$ lines
    of code holding steady for the last six months (Figure~\ref{fig:coverage}).
    Some of the historical components of the code base would still benefit from
    increased test coverage. Documentation for the code is automatically built and hosted 
    by the CircleCI service to facilitate evaluation of documentation changes / integrity.
    Our full test suite also passes with PyPy3, a just-in-time compiled version
    of the Python language.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/coverage-chart}
\caption{\% coverage of the SciPy code base by unit tests over the six
	 months preceding June 28, 2018 as reported by the \texttt{codecov}
	 service. The analysis of lines of source code covered is performed
	 automatically in our continuous integation suite using the
	 \texttt{pytest-cov} and \texttt{gcov} libraries.}
\label{fig:coverage}
\end{figure}

    % NOTE: is there a citation for PyPy?

@@

The SciPy Roadmap\cite{SciPy_roadmap} is a continuously-updated document
maintained by the community that describes some of the major directions
for improvement for the project, as well as specific limitations and
matters that require assistance moving forward.

We are still striving to increase the number of SciPy usage tutorials beyond
our current 15 section offering\cite{SciPy_tutorials}, but our
standard documentation for library features is already in excellent shape.

The low-level Cython code in our library (which interacts with C-level
code and exposes it for Python usage) could use some measure of modernization,
including migration to typed memoryviews to handle NumPy arrays instead of
the older syntax:

\begin{lstlisting}[language=Python]
# example of a function that takes a single
# 3 dimensional typed memoryview as an argument
# this allows Cython-level handling of a NumPy
# array object
cpdef int func_new(int[:, :, :] arr):
    pass

# example of the old syntax for handling
# the same scenario
cpdef int func_old(object[int, ndim=3, mode='strided'] arr):
    pass

\end{lstlisting}

\texttt{fftpack} and {linalg} have too much functionality overlap with
\texttt{numpy} equivalents.

\texttt{scipy.interpolate} could benefit from a set of new features including
spline fitting routines with better user control; arithmetic routines for splines;
transparent tensor-product splines, Non-Uniform Rational B-Splines support;
and mesh refinement and coarsening of B-splines and corresponding tensor products.
Some spline functionality in \texttt{scipy.signal} should be migrated to \texttt{interpolate} 
and Second Order Sections need to be updated to match the capabilities in other routines.

For \texttt{scipy.linalg}, the major change needed is to support a more recent
version of LAPACK. Users should be able access an increasing repertoire of
LAPACK functions using either \texttt{get\_lapack\_funcs} or
\texttt{sp.linalg.lapack.name\_of\_function}.

\texttt{scipy.optimize} would benefit from a few more good global optimizers 
as well as large-scale optimizers.

\texttt{scipy.ndimage} should be completely migrated to a data point
model (values on a grid), instead of a pixel model (elements with centers),
both of which have been used in the past, because the data point model is now
proving more effective in practice.

While \texttt{scipy.sparse} is mature, there are many data
formats and the community continues to move forward here so we will aim
to take advantage of their improvements\cite{Sparse_Arrays} 
while preserving backwards compatibility to the extent that it is possible.
\texttt{sparse.linalg} may benefit from wrappers for PROPACK 
for faster sparse SVD computation.

\texttt{scipy.spatial} will benefit from \texttt{spatial.transforms} with
support for rotation matrices.

\texttt{scipy.special} needs precision improvements for hypergeometric 
functions, parabolic cylinder functions, and spheroidal wave functions.
